---
title: "Final Individual Report"
author: "Lauren M. Rich"
date: "2026-05-21"
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 3
    embed-resources: true
    code-fold: true
    code-summary: "Show Code"
execute:
  warning: false
  message: false
---

# Livestock Methane Emissions in the United States (1990–2024)

## 1. Introduction
Greenhouse gas emissions from human activity are one of the key drivers of US-based (and Global) climate change. While carbon dioxide ($CO_2$) makes up the largest total volume of these gases, methane ($CH_4$) is a key target if we want to slow down warming. While Methane does not stay in the atmosphere as long as carbon dioxide (about 12 years compared to centuries for $CO_2$) it traps heat much more effectively.

To measure this impact, a metric called Global Warming Potential ($GWP$) is used. Over a 100-year timeline ($GWP_{100}$), one ton of methane causes about 28 to 36 times more warming than a ton of carbon dioxide. If we shorten that timeline to  20 years ($GWP_{20}$), methane is 80 to 86 times more powerful than $CO_2$. As methane is so potent in the short term, cutting methane emissions quickly can have an immediate cooling effect on the planet.

The agricultural sector, particularly livestock production, is one of the largest sources of methane emissions. This report focuses on two main questions to help guide environmental policy:
1. **What percentage of total US methane emissions comes from agriculture?**
2. **What does the breakdown of those emissions look like when we look at individual animal species?**

This report establishes an accurate baseline of livestock emissions in the US from 1990 to 2024.

---

## 2. Data Acquisition and Processing

### 2.1 Data Ingestion and API Setup
The data pipeline is split into two main parts: tracking overall national sectors and breaking down emissions by animal species.

For the macro sector view, data was pulled directly from the Climate Watch historical emissions database. Climate Watch collects national greenhouse gas inventories put together by the World Resources Institute (WRI). We used the `httr2` package in R to connect to the Climate Watch API endpoint:
`https://www.climatewatchdata.org/api/v1/data/historical_emissions`

The code filters the request specifically for the United States (`regions[] = "USA"`), focuses exclusively on methane gas (`gases[] = 554`), and covers the years from 1990 to 2024.

For the species-level breakdown, data was pulled from the United Nations Food and Agriculture Organization (FAOSTAT), using the Global Livestock Environmental Assessment Model (GLEAM). As the FAOSTAT API limits large requests, the script splits the animal codes into smaller groups of 50. A loop runs through each year, pausing for 0.1 seconds (`Sys.sleep(0.1)`) between calls so the connection stays stable. The script then combines these pieces into a single dataset using `bind_rows()` and saves it locally in `data/group_project/faostat/`.

### 2.2 Data Cleaning and Formatting
A challenign part of parts of pulling data from APIs is handling nested formats like JSON. The Climate Watch API returns a clean table for the metadata, but the actual annual emissions are buried inside a nested list of dataframes within a single column named `emissions`. The R script uses `tidyr::unnest()` to flatten this structure so we can work with it.

To prevent errors if the API changes in the future, the script includes a safety check:
$$\Delta_{\text{cols}} = \text{names}(cw\_long) \setminus \text{names}(cw\_df)$$

This ensures that only two new columns—the year and the emission value—are pulled out of the nested data. Once verified, the script converts these columns to integers and numbers, and drops any missing rows (`is.na`).

### 2.3 Sector Normalization
We made one major adjustment to the data: **we removed "Fugitive Emissions" entirely.** Fugitive emissions are accidental leaks, venting, and flaring that happen when fossil fuels are extracted and moved. 

While fugitive emissions are a major environmental issue, including them makes it harder to see the steady baseline emissions from everyday infrastructure like farming or waste management. By focusing only on fixed sectors (Agriculture, Energy, Waste, and Industrial Processes), we get a clearer view.

The percentage share ($P_i$) for any given sector is calculated like this:

$$P_i = \frac{\sum_{t=1990}^{2024} E_{i,t}}{\sum_{j=1}^{n} \left( \sum_{t=1990}^{2024} E_{j,t} \right)} \times 100$$

Here, $E$ is the sum of emissions for a sector, $t$ represents the years, and $j$ loops through our four main sectors. This step ensures our charts reflect stable domestic infrastructure, without being distorted by sudden spikes or drops in oil and gas leaks.

---

## 3. Analysis: The  Footprint of Agriculture

The Climate Watch data reveals a clear picture of methane production in the United States. When looking at the total methane emissions across our isolated sectors over this 34-year period, **Agriculture** stands out as one of the largest contributors.

Agriculture accounts for a  majority of the country's structural methane emissions. The remaining emissions are split between the Waste sector (mostly methane gas escaping from landfills and wastewater treatment), the Energy sector (from burning fuels and utility operations), and a very small amount from heavy Industrial Processes.

This breakdown matters a lot for climate policy. Most  conversations about climate change focus on cutting emissions from energy and transportation. While energy transition is essential for carbon dioxide, this data proves that for methane, the biggest opportunity for improvement is in agriculture. 

Furthermore, agricultural methane is biological, which means that it comes from living beings rather than mechanical failures. This makes it harder to fix. You can fix an energy leak by patching a pipe, but you cannot stop natural biological processes without changing how we manage farms and livestock.

---

## 4. Micro Livestock Analysis: Species-Level

To answer our second question, we isolated the livestock data from the UN FAOSTAT database, focusing on nine primary types of animals: Cattle, Swine, Horses, Sheep, Goats, Chickens, Turkeys, Mules/Asses, and Ducks. 

The results show that livestock emissions are heavily concentrated in one place. Instead of being spread evenly across different types of meat and dairy farming, methane emissions are clearly driven by **Cattle**.

| Animal Species | Data Status | Share of Livestock Methane |
| :--- | :--- | :--- |
| **Cattle** (Beef and Dairy) | Included | ~ 95.0% to 97.0% |
| **Swine** (Pigs) | Included | ~ 1.5% to 2.5% |
| **Sheep & Goats** | Included | < 1.0% |
| **Poultry** (Chickens, Turkeys, Ducks) | Included | < 0.1% |

Cattle—including both beef herds and industrial dairy cows—account for roughly 95.5% to 97.0% of all livestock methane in the United States. Swine production is the next largest source, but it represents a tiny fraction of the total. The remaining animals—such as horses, sheep, goats, and all poultry—have a very small impact on the country's total methane footprint.

As the differences between these animals are so massive, the R code applies a smart formatting rule to the chart legend. For animals with high emissions like cattle and pigs, the percentages are rounded to one decimal place to keep the labels clean. For animals with tiny footprints like ducks or working mules, the code  switches to three decimal places. Without this adjustment, these smaller categories would simply round down to 0.0%, hiding them from the final report.

---

## 5. Drivers of Variance

The huge gap between cattle and other livestock comes down to two things: basic animal biology and how farms manage animal waste.

### 5.1 Enteric Fermentation (Digestive Emissions)
The primary reason cattle dominate methane emissions is a digestive process called **enteric fermentation**. Cattle, sheep, and goats are ruminants, meaning they have a specialized stomach split into multiple compartments, the largest of which is the rumen.

Ruminants rely on a complex mix of microbes (like bacteria and fungi) to break down tough, fibrous plants like grass and hay. During this process, a group of microbes called methanogens take the carbon dioxide and hydrogen produced by fermentation and turns it into methane gas:

$$CO_2 + 4H_2 \rightarrow CH_4 + 2H_2O$$

The animal then releases this gas into the air. Non-ruminant animals like pigs, horses, and chickens do not have a rumen. While some fermentation happens in a pig or horse's large intestine, it produces a fraction of the methane that a cow produces.

### 5.2 Waste Streams and Manure Management
The second reason for the variation is how manure is handled. While pigs and poultry do not produce much digestive methane, their waste can generate a lot when managed under the wrong conditions.

To create methane, manure needs to be kept in wet, warm environments with no oxygen. In the US, large-scale commercial dairy and pig farms often use liquid handling systems, such as deep pits or open manure lagoons. These liquid systems create an oxygen-free environment where methane-producing bacteria thrive.

On the other hand, poultry farms usually handle waste as a dry material, using dry litter or bedding. Because these dry systems are exposed to oxygen and kept relatively dry, the manure breaks down through aerobic decomposition. This process creates carbon dioxide ($CO_2$) instead of methane. This difference explains why billions of chickens produce almost no measurable methane compared to US cattle herds.

### 5.3 Historical Trends (1990–2024)
Looking at the data over time from 1990 to 2024, the percentage shares have remained  stable. However, underneath that stability, the way American livestock farming operates has changed dramatically. 

In 1990, the US livestock industry was spread across a larger number of smaller farms. Over the last three decades, the industry consolidated into fewer but larger facilities. This shift changed the balance between digestive emissions and manure emissions. 

Improvements in breeding, and high-energy grain diets made individual animals much more efficient. Cows today produce more milk and meat while creating less enteric methane per pound of food. On the other hand, crowding thousands of animals into centralized facilities meant farms relied more heavily on liquid flush systems and large manure lagoons. 

Essentially, the emissions we saved through better animal diets were offset by increased emissions from liquid manure storage. This balance explains why agriculture's share of total methane emissions has remained  at the top of the charts for over 30 years.

---

## 6. Limitations and Data Constraints

While this data pipeline is completely reproducible, the numbers have some limitations.

### 6.1 The "Production Style" Limitation
As we looked into the FAOSTAT and Climate Watch data, we found a major limitation: **the data does not track individual production styles.** The emission numbers for "Cattle" or "Swine" are national totals which means they pool all farming styles together into broad categories.

Because of this, the data cannot separate the impact of different farming models, such as:
* Grass-fed, pasture-raised cattle.
* High-density industrial feedlots (CAFOs).
* Small-scale organic dairy farms.

This makes it difficult to pinpoint the exact footprint of "factory farming" versus traditional farming. For example, a grass-fed cow lives longer and eats a high-fiber diet, meaning it creates more digestive methane over its lifetime. Meanwhile, a feedlot animal eats a grain-heavy diet that reduces its daily digestive emissions, but its concentrated liquid manure creates more lagoon emissions. Because the data bundles these styles together, this report must be viewed as an analysis of the *entire* livestock sector rather than just industrial facilities.

### 6.2 Missing Variables
To make this model more accurate, future research would need to include three missing pieces of information:

* **Weather and Location Data:** Methane production changes based on temperature and humidity, which alter how fast manure breaks down. Mapping livestock locations to regional weather patterns would make national estimates much more accurate.
* **Detailed Feed Data:** The exact mix of fiber and starch in animal feed changes depending on the season and crop prices. Tracking feed supplies over time would help us better estimate changes in digestive emissions.
* **Technology Adoption Rates:** Technologies like anaerobic digesters (which trap gas from manure lagoons) and methane-reducing feed additives alter actual emissions. Because these technologies are often missing from macro databases, our current models might overstate emissions for highly modernized farms.

---

## 7. Conclusions and Policy Recommendations

The data from this  research project offers clear insights for environmental policy. The data confirms that agriculture is the main driver of structural US methane emissions, and  shows that livestock emissions are a reflection of the cattle industry.

For policymakers, this concentration is actually good news because it simplifies the target. We do not need to spread resources thinly across every type of farming; instead, we can focus directly on cattle management. Effective policies should focus on two areas:

* **Diet and Additives:** We need to accelerate the use of methane-reducing feed additives. Adding specific types of seaweed or natural supplements to dairy and feedlot diets can disrupt methane production in the rumen, lowering emissions without hurting animal health.
* **Manure Upgrades:** We need better incentives for farms to move away from open liquid lagoons toward closed anaerobic digesters, especially on large dairy and pig farms. Capturing this gas prevents it from entering the atmosphere and turns it into clean electricity or renewable natural gas.
* **Diet Changes:** We can advocate that individuals start incorporating more plant-based options into their diets, which would help to solve both the methane and animal welfare crisis.


```{R}

#CLIMATEWATCH DATA

#READ IN THE DATA

library(httr2)
library(jsonlite)
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(scales)

# ----------------------------
# 0) Folders
# ----------------------------
data_dir_group_project <- file.path("data", "group_project")
if (!dir.exists(data_dir_group_project)) dir.create(data_dir_group_project, recursive = TRUE)

data_dir_climate_watch <- file.path(data_dir_group_project, "climate_watch")
if (!dir.exists(data_dir_climate_watch)) dir.create(data_dir_climate_watch, recursive = TRUE)

# ----------------------------
# 1) Helper to save JSON responses (your existing approach)
# ----------------------------
save_json_response <- function(url, file_name) {
  out_path <- file.path(data_dir_climate_watch, file_name)
  resp <- request(url) |> req_perform()
  writeBin(resp_body_raw(resp), out_path)
  out_path
}

# ----------------------------
# 2) Pull metadata: data sources, gases, sectors
#    (These endpoints are part of the Climate Watch API /api/v1 structure) [1](https://climatewatch-vizzuality.github.io/climate-watch/_docs/dev/API)
# ----------------------------
data_sources_file <- save_json_response(
  "https://www.climatewatchdata.org/api/v1/data/historical_emissions/data_sources",
  "climate_watch_data_sources.json"
)

gases_file <- save_json_response(
  "https://www.climatewatchdata.org/api/v1/data/historical_emissions/gases",
  "climate_watch_gases.json"
)

sectors_file <- save_json_response(
  "https://www.climatewatchdata.org/api/v1/data/historical_emissions/sectors",
  "climate_watch_sectors.json"
)

data_sources <- fromJSON(data_sources_file, flatten = TRUE)$data
gases <- fromJSON(gases_file, flatten = TRUE)$data
sectors <- fromJSON(sectors_file, flatten = TRUE)$data

# Optional: inspect metadata
# print(data_sources)
# print(gases)
# print(sectors)

# ----------------------------
# 3) IDs (keep yours)
# ----------------------------
source_id <- 274
gas_id <- 554          # you used this for CH4
# sector_id <- 3212     # your old single-sector filter (not used now)

# Pull ALL sectors so bars can be color-coded by sector
# (sectors dataframe typically contains id + name; use the id column)
sector_ids <- sectors$id

# ----------------------------
# 4) Historical emissions request (ALL sectors, USA, 1990–2024)
#    Parameters are passed as arrays using explode, matching your style. [1](https://climatewatch-vizzuality.github.io/climate-watch/_docs/dev/API)
# ----------------------------
historical_req <- request("https://www.climatewatchdata.org/api/v1/data/historical_emissions") |>
  req_url_query(
    `data_sources[]` = source_id,
    `gases[]`        = gas_id,
    `sectors[]`      = sector_ids,   # <-- ALL sectors
    `regions[]`      = "USA",
    start_year       = 1990,
    end_year         = 2025,
    .multi           = "explode"
  )

historical_file <- file.path(
  data_dir_climate_watch,
  "climate_watch_historical_emissions_usa_ch4_all_sectors_1990_2024.json"
)

historical_resp <- historical_req |> req_perform()
writeBin(resp_body_raw(historical_resp), historical_file)

# ----------------------------
# 5) Load JSON -> data frame (this will include nested emissions)
# ----------------------------
climate_watch_raw <- fromJSON(historical_file, flatten = TRUE)
cw_df <- as_tibble(climate_watch_raw$data)

# ----------------------------
# 6) Unnest emissions (THIS is where the years/values live)
#    Your emissions column contains <data.frame[34 x 2]> per row.
# ----------------------------
cw_long <- cw_df %>%
  unnest(emissions)

# Identify the two new columns that came from emissions (usually year + value)
new_cols <- setdiff(names(cw_long), names(cw_df))
if (length(new_cols) != 2) {
  stop(
    "Expected exactly 2 columns inside `emissions`, but found: ",
    paste(new_cols, collapse = ", "),
    "\nRun: names(cw_df$emissions[[1]]) to see the nested column names."
  )
}

cw_long <- cw_long %>%
  rename(
    year_raw  = !!sym(new_cols[1]),
    value_raw = !!sym(new_cols[2])
  ) %>%
  mutate(
    year  = suppressWarnings(as.integer(year_raw)),
    value = suppressWarnings(as.numeric(value_raw))
  ) %>%
  filter(!is.na(year), !is.na(value))

# --- INSERT ESTIMATION HERE ---
# 6.5) Create 2025 projection
# This clones 2024 data and adjusts it.
projected_2025 <- cw_long %>% 
  filter(year == 2024) %>%
  mutate(
    year = 2025,
    # Example: Assume 0.5% decrease (change to suit your research)
    value = value * 0.995, 
    is_estimated = TRUE
  )

# Add the 2025 data to the main dataset
cw_long <- bind_rows(cw_long, projected_2025)
# ------------------------------


# ----------------------------
# 7) Filter to CH4 (your goal) + 5-year bins
#    (Your dataset may still contain multiple gases; filter here.)
# ----------------------------
cw_ch4 <- cw_long %>%
  filter(gas == "CH4") %>%
  filter(year >= 1990, year <= 2024) %>%
  mutate(
    period_start = year - (year %% 5),
    period_end   = period_start + 4L,
    period       = paste0(period_start, "–", period_end)
  )

```

```{R}
## CLIMATE WATCH CHARTS

library(stringr)

## CLIMATE WATCH CHARTS

# 1) Clean and Aggregate over 1990–2024
cw_total <- cw_ch4 %>%
  filter(year >= 1990, year <= 2024) %>% 
  # 1. Strip any accidental spaces from the sector names
  mutate(sector = trimws(as.character(sector))) %>%
  # 2. Drop anything containing "Total" or "Fugitive" safely
  filter(
    !str_detect(sector, regex("Total", ignore_case = TRUE)),
    !str_detect(sector, regex("Fugitive", ignore_case = TRUE))
  ) %>% 
  group_by(sector) %>%
  summarise(
    total_emissions = sum(value, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    percentage = total_emissions / sum(total_emissions),
    label = percent(percentage, accuracy = 0.1)
  )

# 2) Plot the Donut Chart
p <- ggplot(cw_total, aes(x = 2, y = percentage, fill = reorder(sector, -percentage))) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +  
  geom_text(aes(label = label), 
            position = position_stack(vjust = 0.5), 
            size = 3.5, fontface = "bold") +
  theme_void() + 
  theme(
    legend.position = "right",
    plot.title = element_text(face = "bold", hjust = 0.5, size = 14)
  ) +
  labs(
    title = "USA CH4 Emissions: Sector Composition",
    fill  = "Sector"
  )

print(p)

```



```{r}

## FAOSTAT DATA

# 1) Load Libraries -------------------------------------------
library(dplyr)
library(ggplot2)

# 2) Load the Data you saved today ----------------------------
# This uses the specific location you created
file_path <- file.path("data", "group_project", "faostat", "methane_us_1990plus.csv")

if(file.exists(file_path)) {
  methane_us <- read.csv(file_path)
  message("Data loaded successfully!")
} else {
  stop("File not found! Make sure you are in the correct R Project.")
}


```

```{r}
# 1) Define the exact categories
clean_list <- c("Cattle", "Swine", "Horses", "Sheep", "Goats", 
                "Chickens", "Turkeys", "Mules and Asses", "Ducks")

# 2) Filter and apply conditional formatting for decimals
donut_data_custom <- methane_us %>%
  filter(Year >= 1990) %>% 
  filter(grepl("CH4", Element)) %>% 
  filter(Item %in% clean_list) %>%
  group_by(Item) %>%
  summarise(Total_Value = sum(Value, na.rm = TRUE), .groups = 'drop') %>%
  mutate(
    perc_num = (Total_Value / sum(Total_Value)) * 100,
    # Apply 3 decimals ONLY to Mules/Asses and Ducks; 1 decimal for everyone else
    label = ifelse(Item %in% c("Mules and Asses", "Ducks"),
                   paste0(Item, " (", sprintf("%0.3f", perc_num), "%)"),
                   paste0(Item, " (", sprintf("%0.1f", perc_num), "%)"))
  ) %>%
  arrange(desc(Total_Value))

# 3) Create the Donut Chart
ggplot(donut_data_custom, aes(x = 2, y = Total_Value, fill = reorder(label, -Total_Value))) +
  geom_col(color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void() +
  labs(
    title = "Livestock Methane Comparison",
    fill = "Animal Species (%)"
  ) +
  theme(
    plot.title = element_text(face = "bold", size = 16, hjust = 0.5),
    legend.position = "right",
    legend.text = element_text(size = 10)
  )

```

```{r faostat_scrape}
# =============================================================
# 1) DATA ACQUISITION: WEB SCRAPE AND SAVE LOCALLY (DON'T RUN)
# =============================================================

# DECISIVE BYPASS: This stops the code from running even if Quarto ignores chunk settings
if (TRUE) {
  message("Skipping live API scrape. Using cached data folder resources.")
} else {
  # The rest of your code is safely locked away in this 'else' block and cannot run
  
  # Setup Directories  
  data_dir_faostat <- file.path("data", "group_project", "faostat")
  if(!dir.exists(data_dir_faostat)) dir.create(data_dir_faostat, recursive = TRUE)

  # Authentication Environment Variable
  Sys.setenv(FAOSTAT_TOKEN = "eyJraWQiOiJVSFE2dmw...")

  library(httr)
  library(stringr)
  library(tibble)
  library(jsonlite)

  base_url <- "[https://faostatservices.fao.org/api/v1](https://faostatservices.fao.org/api/v1)"

  fao_get_json <- function(path, query = list()) {
    resp <- GET(paste0(base_url, path), query = query, add_headers(Authorization = paste("Bearer", Sys.getenv("FAOSTAT_TOKEN"))))
    stop_for_status(resp)
    js <- fromJSON(content(resp, as = "text", encoding = "UTF-8"), flatten = TRUE)
    as_tibble(js$data)
  }

  fao_get_csv <- function(path, query = list()) {
    resp <- GET(paste0(base_url, path), query = query, add_headers(Authorization = paste("Bearer", Sys.getenv("FAOSTAT_TOKEN"))))
    stop_for_status(resp)
    read.csv(text = content(resp, as = "text", encoding = "UTF-8"), stringsAsFactors = FALSE, colClasses = "character", na.strings = c("", "NA")) %>% as_tibble()
  }

  # Metadata & Filtering Logic
  area_meta <- fao_get_json("/en/definitions/domain/GLE/area") %>% rename(code = `Country Code`, label = `Country`)
  elem_meta <- fao_get_json("/en/definitions/domain/GLE/element") %>% rename(code = `Element Code`, label = `Element`)
  item_meta <- fao_get_json("/en/definitions/domain/GLE/item") %>% rename(code = `Item Code`, label = `Item`)

  us_code <- area_meta %>% filter(str_detect(label, regex("^United States", ignore_case = TRUE))) %>% slice(1) %>% pull(code)
  ch4_codes_raw <- elem_meta %>% filter(str_detect(label, regex("methane|\\bch4\\b", ignore_case = TRUE))) %>% pull(code)
  ch4_codes_string <- paste(ch4_codes_raw, collapse = ",")

  # Multi-Year Extraction Loop
  years <- 1990:2024
  item_chunks <- split(as.character(item_meta$code), ceiling(seq_along(item_meta$code) / 50))

  results_list <- lapply(years, function(y) {
    Sys.sleep(0.1) 
    bind_rows(lapply(item_chunks, function(items) {
      fao_get_csv("/en/data/GLE", query = list(
        area = us_code, 
        element = ch4_codes_string, 
        item = paste(items, collapse = ","), 
        year = y, 
        datasource = "DB4", 
        output_type = "csv"
      ))
    }))
  })

  methane_us_1990plus <- bind_rows(results_list) %>%
    mutate(Year = as.integer(Year), Value = as.numeric(Value))

  # Export Data to Local Cache File
  file_destination <- file.path(data_dir_faostat, "methane_us_1990plus.csv")
  write.csv(methane_us_1990plus, file = file_destination, row.names = FALSE)
}