---
title: "Mini-Project 02 - How Do You Do ‘You Do You’?"
author: "laurenrichbetch"
format:
  html:
    code-fold: false
    code-tools: false
execute:
  echo: false
  warning: false
  message: false
editor:
  mode: source
---

## Introduction

In this mini project, I will analyze how people spend their time.

## Load Data
```{r}
library(tidyverse)
library(glue)
library(readxl)
library(httr2)

## Filter Data

load_atus_data <- function(file_base = c("resp", "rost", "act")){
  if(!dir.exists(file.path("data", "mp02"))){
    dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
  }
  
  file_base <- match.arg(file_base)
  
  file_name_out <- file.path("data", 
                             "mp02", 
                             glue("atus{file_base}_0324.dat"))
  
  if(!file.exists(file_name_out)){
    
    url_end <- glue("atus{file_base}-0324.zip")
    
    temp_zip <- tempfile(fileext=".zip")
    temp_dir <- tools::file_path_sans_ext(temp_zip)
    
    request("https://www.bls.gov") |>
      req_url_path("tus", "datafiles", url_end) |>
      req_headers(`User-Agent` = "Mozilla/5.0 (Macintosh; Intel Mac OS X 10.15; rv:143.0) Gecko/20100101 Firefox/143.0") |>
      req_perform(temp_zip)
    
    unzip(temp_zip, exdir = temp_dir)
    
    temp_file <- file.path(temp_dir, basename(file_name_out))
    
    file.copy(temp_file, file_name_out)
  }
  
  file <- read_csv(file_name_out, 
                   show_col_types = FALSE)
  
  switch(file_base,
         resp = file |> 
           rename(survey_weight = TUFNWGTP, 
                  time_alone = TRTALONE,
                  hourly_wage = TEERNH1O,
                  survey_year = TUYEAR,
                  num_children = TRCHILDNUM) |>
           mutate(diary_month_num = month(ymd(TUDIARYDATE)),  # ADD: Month number (1-12)
                  diary_month = month(ymd(TUDIARYDATE), label = TRUE),  # ADD: Month name
             marital_status = case_match(TRSPPRES, 
                                              1 ~ "Married", 
                                              2 ~ "Married", 
                                              3 ~ "Unmarried",
                                              .default = "Unknown"),
                  employment_status = case_match(TELFS,
                                                 1:2 ~ "Employed",
                                                 3:4 ~ "Unemployed",
                                                 5 ~ "Not in Labor Force",
                                                 .default = "Unknown"),
                  # Boolean for College/University enrollment
                  is_college_student = (TESCHENR == 1 & TESCHLVL == 2)),
         
         rost = file |> 
  mutate(
    sex = case_match(TESEX, 1 ~ "M", 2 ~ "F"),
    age = TEAGE
  ),
         
         act = file |> 
           mutate(activity_n_of_day = TUACTIVITY_N, 
                  time_spent_min = TUACTDUR24, 
                  start_time = TUSTARTTIM, 
                  stop_time = TUSTOPTIME,
                  level1_code = paste0(TRTIER1P, "0000"), 
                  level2_code = paste0(TRTIER2P, "00"), 
                  level3_code = TRCODEP,
                  # Mapping Location codes
                  location = case_match(TEWHERE,
                                        1 ~ "Home",
                                        2 ~ "Workplace",
                                        3 ~ "Someone else's home",
                                        4 ~ "Restaurant/Bar",
                                        5 ~ "Place of worship",
                                        6 ~ "Grocery store",
                                        12 ~ "Vehicle",
                                        .default = "Other/Unknown"))
  ) |> 
    rename(participant_id = TUCASEID) |> 
    select(matches("[:lower:]", ignore.case = FALSE))
}



load_atus_activities <- function(){
  if(!dir.exists(file.path("data", "mp02"))){
    dir.create(file.path("data", "mp02"), showWarnings=FALSE, recursive=TRUE)
  }
  
  dest_file <- file.path("data",
                         "mp02",
                         "atus_activity_codes.csv")
  
  if(!file.exists(dest_file)){
    download.file("https://michael-weylandt.com/STA9750/mini/atus_activity_codes.csv", 
                  quiet=TRUE, 
                  destfile=dest_file, 
                  mode="wb")
  }
  
  read_csv(dest_file, show_col_types = FALSE) 
}


## Create Data Frame

resp <- load_atus_data("resp")
rost <- load_atus_data("rost")
act  <- load_atus_data("act")

## Create Data Frame - Second

activities <- load_atus_activities()

## Glance at data

head(resp, 10)
head(rost, 10)
head(act, 10)
head(activities, 10)
```

## Join data into one combined dataset
```{r}
full_data <- act %>%
  left_join(resp, by = "participant_id") %>%
  left_join(rost, by = "participant_id") %>%
  left_join(activities, by = c("level3_code" = "code_level3"))

```

## How many total hours were recorded on sleeping-type activities
```{r}
total_sleep_hours <- full_data %>%
  filter(level1_code == "010000") %>%
  summarise(total_hours = sum(time_spent_min, na.rm = TRUE) / 60)

total_sleep_hours

## How many female participants reported spending no time alone in 2003?

female_no_alone_2003 <- full_data %>%
  distinct(participant_id, sex, time_alone, survey_year) %>% 
  filter(
    sex == "F",
    survey_year == 2003,
    time_alone == 0
  ) %>%
  summarise(n = n())

female_no_alone_2003

head(full_data, 10)
colnames(full_data)
```

#Weights
```{r}
##Weighted average time alone (all years)

resp %>%
  summarize(avg_time_alone = weighted.mean(time_alone, survey_weight))

##Weighted average time alone (by year)

resp %>%
  group_by(survey_year) %>%
  summarize(avg_time_alone = weighted.mean(time_alone, survey_weight))

##Weighted average time alone (by sex)

full_data %>%
  distinct(participant_id, sex, time_alone, survey_weight) %>%
  group_by(sex) %>%
  summarize(avg_time_alone = weighted.mean(time_alone, survey_weight))

##Weighted count of female participants with zero time alone in 2003

full_data %>%
  distinct(participant_id, sex, time_alone, survey_year, survey_weight) %>%
  filter(sex == "F", survey_year == 2003, time_alone == 0) %>%
  summarize(weighted_count = sum(survey_weight))

##Weighted average number of children under 18

full_data %>%
  distinct(participant_id, num_children, survey_weight) %>%
  summarize(avg_children = weighted.mean(num_children, survey_weight))

##Weighted total time spent in each activity category - In minutes

full_data %>%
  group_by(task_level1) %>%
  summarize(weighted_minutes = sum(time_spent_min * survey_weight, na.rm = TRUE)) %>%
  mutate(weighted_hours = weighted_minutes / 60)
```

## Activity 4
```{r}
library(tidyverse)

##remove duplicate values and create diary_id
full_data <- full_data %>%
  distinct() %>%                                   # remove duplicated rows
  arrange(participant_id, survey_year, activity_n_of_day) %>%
  mutate(diary_id = cumsum(activity_n_of_day == 1))  # new diary starts at activity 1

##Number of respondents
n_respondents <- resp %>%
  distinct(participant_id) %>%
  nrow()
  
## Number of sports-related Level 3 tasks
n_sports <- full_data %>%
  filter(str_detect(task_level3, regex("sport", ignore_case = TRUE))) %>%
  distinct(task_level3) %>%
  nrow()
  
## Percent retired
resp <- resp %>%
  mutate(is_retired = employment_status == "Not in Labor Force")

pct_retired <- resp %>%
  summarize(p = weighted.mean(is_retired, survey_weight, na.rm = TRUE)) %>%
  pull(p) * 100
  
## average hours of sleep per night
sleep_hours <- full_data %>%
  filter(level3_code == "010101") %>%  # Sleeping
  group_by(diary_id, survey_weight) %>%
  summarize(total_sleep = sum(time_spent_min), .groups = "drop") %>%
  summarize(avg_sleep = weighted.mean(total_sleep, survey_weight)) %>%
  pull(avg_sleep) / 60
  
## Average childcare hours per day
childcare_hours <- full_data %>%
  filter(num_children > 0, str_starts(level1_code, "030")) %>%  # Child care category
  group_by(diary_id, survey_weight) %>%
  summarize(total_childcare = sum(time_spent_min), .groups = "drop") %>%
  summarize(avg_childcare = weighted.mean(total_childcare, survey_weight)) %>%
  pull(avg_childcare) / 60

## Diary totals QA
full_data %>%
  group_by(diary_id) %>%
  summarize(total_minutes = sum(time_spent_min)) %>%
  arrange(desc(total_minutes)) %>%
  head()

```



## Activity 4 - Sentences

Since 2003 a total of `r n_respondents` different respondents have completed an ATUS survey.

Americans watch many different sports and ATUS includes `r n_sports` distinct Level‑3 sports‑watching activities.

Approximately `r round(pct_retired, 1)`% of Americans are retired based on a weighted estimate using employment status.

On average Americans sleep about `r round(sleep_hours, 2)` hours per night.

Among Americans with one or more children in the home the average amount of time spent caring for their children is `r round(childcare_hours, 2)` hours per day.


## Activity 5 - Data Tables

```{r}

library(gt)
library(tidyverse)

names(resp)

## Do retired persons spend more time on lawn‑care than full‑time workers?

lawn_table <- full_data %>%
  filter(level3_code == "020901") %>%   # Lawn & garden care
  group_by(diary_id, employment_status, survey_weight) %>%
  summarize(total_lawn = sum(time_spent_min), .groups = "drop") %>%
  mutate(hours = total_lawn / 60) %>%
  group_by(employment_status) %>%
  summarize(avg_hours = weighted.mean(hours, survey_weight)) %>%
  arrange(desc(avg_hours)) %>%
  gt() %>%
  tab_header(title = "Average Daily Lawn-Care Time by Employment Status") %>%
  cols_label(
    employment_status = "Employment Status",
    avg_hours = "Avg Hours per Day"
  ) %>%
  fmt_number(columns = avg_hours, decimals = 2)

lawn_table

## Social & recreational time: married vs unmarried

social_table <- full_data %>%
  filter(str_starts(level1_code, "12")) %>%   # Social & leisure
  group_by(diary_id, marital_status, survey_weight) %>%
  summarize(total_social = sum(time_spent_min), .groups = "drop") %>%
  mutate(hours = total_social / 60) %>%
  group_by(marital_status) %>%
  summarize(avg_hours = weighted.mean(hours, survey_weight)) %>%
  gt() %>%
  tab_header(title = "Social & Recreational Time by Marital Status") %>%
  cols_label(
    marital_status = "Marital Status",
    avg_hours = "Avg Hours per Day"
  ) %>%
  fmt_number(columns = avg_hours, decimals = 2)

social_table

## Where do high‑earning Americans spend the majority of their time?

# 1. Define high earners (top 20% of hourly wage)
full_data <- full_data %>%
  mutate(
    high_earner = hourly_wage >= quantile(hourly_wage, 0.80, na.rm = TRUE)
  )

# 2. Build table of where high earners spend their time
high_earner_table <- full_data %>%
  filter(high_earner == TRUE) %>%
  group_by(diary_id, task_level1, survey_weight) %>%
  summarize(total = sum(time_spent_min), .groups = "drop") %>%
  group_by(task_level1) %>%
  summarize(avg_minutes = weighted.mean(total, survey_weight)) %>%
  arrange(desc(avg_minutes)) %>%
  slice_head(n = 10) %>%
  mutate(hours = avg_minutes / 60) %>%
  gt() %>%
  tab_header(title = "Where High-Earning Americans Spend Their Time") %>%
  cols_label(
    task_level1 = "Activity Category",
    avg_minutes = "Avg Minutes per Day",
    hours = "Avg Hours per Day"
  ) %>%
  fmt_number(columns = c(avg_minutes, hours), decimals = 1)

high_earner_table

## Most Common Activities on Public Transit

library(gt)
library(tidyverse)

transit_table <- full_data %>%
  filter(location == "Vehicle") %>%
  count(task_level3, sort = TRUE) %>%
  mutate(percent = n / sum(n) * 100) %>%   # compute percent BEFORE slicing
  slice_head(n = 10) %>%
  select(task_level3, percent) %>%         # keep only the columns you want
  gt() %>%
  tab_header(title = "Most Common Activities While in a Vehicle") %>%
  cols_label(
    task_level3 = "Activity",
    percent = "% of Total"
  ) %>%
  fmt_number(columns = percent, decimals = 1)

transit_table

```

#View Activities
```{r}
full_data %>%
  distinct(task_level1) %>%
  arrange(task_level1)

```

#View Activities
```{r}
names(full_data)
```

## Activity 5 - Data Tables - Cont'd

```{r}
## Create age groups
full_data <- full_data %>%
  mutate(
    age_group = case_when(
      age >= 18 & age <= 29 ~ "18–29",
      age >= 30 & age <= 44 ~ "30–44",
      age >= 45 & age <= 59 ~ "45–59",
      age >= 60 ~ "60+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group))   # <-- remove NA age groups


## Filter to my fav activities
favorite_activities <- c(
  "Eating and Drinking",
  "Traveling",
  "Sports, Exercise, and Recreation",
  "Personal Care"
)


## Compute average minutes per day by age group
favorite_table <- full_data %>%
  filter(task_level1 %in% favorite_activities) %>%
  group_by(diary_id, age_group, task_level1, survey_weight) %>%
  summarize(total_minutes = sum(time_spent_min), .groups = "drop") %>%
  group_by(age_group, task_level1) %>%
  summarize(avg_minutes = weighted.mean(total_minutes, survey_weight), .groups = "drop") %>%
  mutate(hours = avg_minutes / 60) %>%     # <-- add hours column
  arrange(task_level1, age_group)


## Table Build
favorite_table %>%
  mutate(
    avg_minutes = round(avg_minutes, 1),
    hours = round(hours, 2)                # <-- format hours
  ) %>%
  gt() %>%
  tab_header(title = "Time Spent per Day on Favorite Activities by Age Group") %>%
  cols_label(
    age_group = "Age Group",
    task_level1 = "Activity",
    avg_minutes = "Avg Minutes per Day",
    hours = "Avg Hours per Day"
  ) %>%
  fmt_number(columns = c(avg_minutes, hours), decimals = 1)

```

## Activity 6 - Graph #1

```{r}
##  How does the amount of time spent on household activities change as Americans age?

full_data <- full_data %>%
  mutate(
    age_group = case_when(
      age >= 18 & age <= 29 ~ "18–29",
      age >= 30 & age <= 44 ~ "30–44",
      age >= 45 & age <= 59 ~ "45–59",
      age >= 60 ~ "60+",
      TRUE ~ NA_character_
    )
  ) %>%
  filter(!is.na(age_group))


household_plot_data <- full_data %>%
  filter(task_level1 == "Household Activities") %>%
  group_by(diary_id, age_group, survey_weight) %>%
  summarize(total_minutes = sum(time_spent_min), .groups = "drop") %>%
  group_by(age_group) %>%
  summarize(avg_minutes = weighted.mean(total_minutes, survey_weight), .groups = "drop")


library(ggplot2)

ggplot(household_plot_data, aes(x = age_group, y = avg_minutes, group = 1)) +
  geom_line(linewidth = 1.3, color = "#003f7f") +     # dark blue line
  geom_point(size = 3.5, color = "#66b2ff") +         # light blue dots
  labs(
    title = "Average Time Spent on Household Activities by Age Group",
    x = "Age Group",
    y = "Minutes per Day"
  ) +
  theme_minimal(base_size = 15)

```
## Write up

As people get older they spend more time on household activities - this makes sense as the older community (60+) is largely going to be retired and spend more time at home versus younger demographics, who are working and out of the house.  It's also possible that the same tasks that used to be completed in a shorter amount of time now take longer as the older demographics do not move as quickly as they once had - they would also need to be mindful of moving in a way that would prevent injuries.


## Activity 6 - Graph #2

```{r}

library(dplyr)
library(ggplot2)

# 1. Identify people who do education-related activities
education_people <- full_data %>%
  filter(task_level1 == "Education") %>%
  pull(participant_id) %>%
  unique()

# 2. Look at ALL their activities during summer months
summer_activities <- full_data %>%
  filter(
    participant_id %in% education_people,
    diary_month_num %in% c(6, 7, 8),
    !is.na(task_level1)
  )

# 3. Count the most common activities
activity_counts <- summer_activities %>%
  count(task_level1, sort = TRUE) %>%
  slice_max(n, n = 10)

# 4. Plot
ggplot(activity_counts, aes(x = reorder(task_level1, n), y = n)) +
  geom_col(fill = "#2C7BB6") +
  coord_flip() +
  labs(
    title = "Most Popular Summer Activities Among Americans Engaged in Education Activities",
    x = "Activity",
    y = "Count"
  ) +
  theme_minimal(base_size = 14)

```

## Write up

During the summer months, people who work in Education spend most of their time on personal care, with Socializing, Relaxing and Leisure as a close second.  This makes sense as working in education during the school year requires early wake up times, and lots of energy – thus, making the summer the perfect opportunity to relax, refresh and reset.


## Activity 6 - Graph #3

```{r}


library(dplyr)
library(ggplot2)


# 1. Identify respondents enrolled in higher education
students <- full_data %>%
  filter(is_college_student == 1)


# 2. Filter to education-related activities
#    (task_level1 == "Education")
student_ed_activities <- students %>%
  filter(task_level1 == "Education")


# 3. Compute total time spent on education per participant
student_time <- student_ed_activities %>%
  group_by(participant_id, age) %>%
  summarise(
    total_ed_minutes = sum(time_spent_min, na.rm = TRUE),
    .groups = "drop"
  )


# 4. Create age groups
student_time <- student_time %>%
  mutate(age_group = case_when(
    age < 25 ~ "18–24",
    age < 35 ~ "25–34",
    age < 45 ~ "35–44",
    age < 55 ~ "45–54",
    TRUE ~ "55+"
  ))


# 5. Compute average time spent per age group
age_summary <- student_time %>%
  group_by(age_group) %>%
  summarise(
    avg_minutes = mean(total_ed_minutes, na.rm = TRUE),
    n = n()
  )


# 6. Plot the results
ggplot(age_summary, aes(x = age_group, y = avg_minutes)) +
  geom_col(fill = "#2C7BB6") +
  labs(
    title = "Average Time Spent on Education Activities by Age Group",
    x = "Age Group",
    y = "Average Minutes per Day"
  ) +
  theme_minimal(base_size = 14)


```

## Write up

The amount of time spent on educational activities is comparable across age groups. 


## Activity 6 - Graph #4

```{r}

library(dplyr)
library(ggplot2)

# 1. Filter to sports-related activities (ATUS code 130000)
sports <- full_data %>%
  filter(code_level1 == "130000") %>%   # SPORTS category
  filter(!is.na(task_level3))           # actual sport names

# 2. Create age groups
sports <- sports %>%
  mutate(age_group = cut(
    age,
    breaks = c(15, 25, 35, 45, 55, 65, 100),
    labels = c("15–24", "25–34", "35–44", "45–54", "55–64", "65+"),
    right = FALSE
  ))

# 3. Compute total minutes per sport per age group
sport_summary <- sports %>%
  group_by(age_group, task_level3) %>%
  summarise(total_minutes = sum(time_spent_min, na.rm = TRUE), .groups = "drop")

# 4. Identify the top 6 sports overall
top_sports_overall <- sport_summary %>%
  group_by(task_level3) %>%
  summarise(total = sum(total_minutes)) %>%
  arrange(desc(total)) %>%
  slice_head(n = 6) %>%
  pull(task_level3)

sport_summary <- sport_summary %>%
  filter(task_level3 %in% top_sports_overall)

# 5. Rank sports within each age group
sport_ranks <- sport_summary %>%
  group_by(age_group) %>%
  mutate(rank = rank(-total_minutes, ties.method = "first")) %>%
  ungroup()

# 6. Bump plot using sport names
ggplot(sport_ranks, aes(
  x = age_group,
  y = rank,
  color = task_level3,
  group = task_level3
)) +
  geom_line(linewidth = 1.5) +
  geom_point(size = 3) +
  scale_y_reverse(breaks = 1:6) +
  labs(
    title = "How Interest in Specific Sports Changes Across Age Groups",
    x = "Age Group",
    y = "Rank (1 = Most Time Spent)",
    color = "Sport"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  )

```

## Write up

It is unsurprising that participation in water sports significantly decreases as people age – water sports require people to be very fit, as it can often mean treading water for extended periods of time – as people age, they lose muscle mass, and therefore, it can be dangerous to partake in a competitive sport in the water.  Likewise, weight-lifting and/or strength-training also skews young, as it can be challenging and unsafe to lift heavy weights for older demographics.  More people play golf as they continue to age as it’s not a strenuous sport.  Walking and fishing trend flat across demos as these are also low-impact sports/ activities.

## Activity 6 - Graph #5

```{r}

library(dplyr)
library(ggplot2)

# 1. Filter to childcare activities
childcare <- full_data %>%
  filter(task_level1 == "Caring for and Helping Household Members") %>%
  filter(!is.na(num_children))   # ensure family size is known

# 2. Compute total childcare minutes per parent per day
parent_time <- childcare %>%
  group_by(participant_id, num_children) %>%
  summarise(
    total_childcare_minutes = sum(time_spent_min, na.rm = TRUE),
    .groups = "drop"
  )

# 3. Compute time per child
parent_time <- parent_time %>%
  mutate(minutes_per_child = total_childcare_minutes / num_children)

# 4. Summarize average time per child by family size
family_summary <- parent_time %>%
  group_by(num_children) %>%
  summarise(
    avg_minutes_per_child = mean(minutes_per_child, na.rm = TRUE),
    n = n()
  )

# 5. Plot: Time per child vs. number of children
ggplot(family_summary, aes(x = num_children, y = avg_minutes_per_child)) +
  geom_line(linewidth = 1.5, color = "#2C7BB6") +
  geom_point(size = 3, color = "#2C7BB6") +
  labs(
    title = "How Parental Time Per Child Changes as Family Size Increases",
    x = "Number of Children in Household",
    y = "Average Minutes of Childcare Per Child"
  ) +
  theme_minimal(base_size = 14)


```
## Write up

Time per child decreases as the number of children in the household increases – this makes so much sense because each parent only has a set amount of hours in the day, and as more children enter the household, there is simply less time to be spent on each one of time.

## Activity 7

```{r}

library(dplyr)
library(ggplot2)

# 0. ACTIVITY MAP (ATUS Level 1 Codes)
activity_map <- list(
  sleep = "010000",
  exercise = "130000",
  social = "120000",
  chores = "020000",
  work = "050000",
  childcare = "030000",
  education = "060000",
  leisure = "120000"
)

# 1. DEMOGRAPHIC FILTERS

# --- Stay-at-home dads ---
dads <- full_data %>%
  filter(sex == "M") %>%
  filter(num_children > 0) %>%
  filter(employment_status == "Not in Labor Force")

# --- College undergraduates ---
college <- full_data %>%
  filter(is_college_student == TRUE)

# --- Sophia demographic ---
sophia <- full_data %>%
  filter(sex == "F") %>%
  filter(age >= 30, age <= 39) %>%
  filter(marital_status == "Unmarried") %>%
  filter(employment_status == "Employed")

# 2. FUNCTION TO PROCESS ANY DEMOGRAPHIC

process_group <- function(df, activities) {
  
  df_daily <- df %>%
    mutate(activity = case_when(
      code_level1 == activity_map$sleep ~ "Sleep",
      code_level1 == activity_map$exercise ~ "Exercise",
      code_level1 == activity_map$social ~ "Socializing",
      code_level1 == activity_map$chores ~ "Chores",
      code_level1 == activity_map$work ~ "Work",
      code_level1 == activity_map$childcare ~ "Childcare",
      code_level1 == activity_map$education ~ "Studying",
      TRUE ~ NA_character_
    )) %>%
    filter(activity %in% activities) %>%
    group_by(participant_id, activity) %>%
    summarise(total_minutes = sum(time_spent_min, na.rm = TRUE), .groups = "drop")
  
  # Typical levels
  typical <- df_daily %>%
    group_by(activity) %>%
    summarise(
      low = quantile(total_minutes, 0.25, na.rm = TRUE),
      median = quantile(total_minutes, 0.50, na.rm = TRUE),
      high = quantile(total_minutes, 0.75, na.rm = TRUE)
    )
  
  list(
    daily = df_daily,
    typical = typical
  )
}

# 3. PROCESS EACH GROUP

# Stay-at-home dads
dads_results <- process_group(
  dads,
  activities = c("Childcare", "Chores", "Sleep", "Socializing", "Exercise")
)

# College undergraduates
college_results <- process_group(
  college,
  activities = c("Sleep", "Studying", "Socializing", "Exercise")
)

# Sophia
sophia_results <- process_group(
  sophia,
  activities = c("Sleep", "Exercise", "Socializing", "Chores", "Work")
)

# 4. TARGET LEVELS FOR EACH GROUP

targets_dads <- tibble(
  activity = c("Childcare", "Chores", "Sleep", "Socializing", "Exercise"),
  target_minutes = c(180, 90, 480, 60, 30)
)

targets_college <- tibble(
  activity = c("Sleep", "Studying", "Socializing", "Exercise"),
  target_minutes = c(480, 180, 60, 30)
)

targets_sophia <- tibble(
  activity = c("Sleep", "Exercise", "Socializing", "Chores", "Work"),
  target_minutes = c(450, 45, 60, 30, 510)
)

# 5. FUNCTION TO COMPUTE % MEETING TARGETS

compute_target_hits <- function(df_daily, targets) {
  df_daily %>%
    left_join(targets, by = "activity") %>%
    mutate(hit = total_minutes >= target_minutes) %>%
    group_by(activity) %>%
    summarise(percent_meeting = mean(hit, na.rm = TRUE))
}

# Compute for each group
dads_hits <- compute_target_hits(dads_results$daily, targets_dads)
college_hits <- compute_target_hits(college_results$daily, targets_college)
sophia_hits <- compute_target_hits(sophia_results$daily, targets_sophia)

# 6. CHARTS FOR EACH GROUP

plot_typical <- function(typical_df, title) {
  ggplot(typical_df, aes(x = activity, y = median)) +
    geom_col(fill = "#4C72B0") +
    labs(title = title, x = "Activity", y = "Median Minutes") +
    theme_minimal()
}

plot_hits <- function(hits_df, title) {
  ggplot(hits_df, aes(x = activity, y = percent_meeting)) +
    geom_col(fill = "#55A868") +
    labs(title = title, x = "Activity", y = "Percent Meeting Target") +
    theme_minimal()
}

# --- Stay-at-home dads ---
plot_typical(dads_results$typical, "Typical Time Use: Stay-at-Home Dads")
plot_hits(dads_hits, "Percent Meeting Targets: Stay-at-Home Dads")

# --- College undergraduates ---
plot_typical(college_results$typical, "Typical Time Use: College Undergraduates")
plot_hits(college_hits, "Percent Meeting Targets: College Undergraduates")

# --- Sophia ---
plot_typical(sophia_results$typical, "Typical Time Use: Sophia Demographic")
plot_hits(sophia_hits, "Percent Meeting Targets: Sophia Demographic")

# 7. PRINT TABLES

print("Stay-at-home dads: Typical Levels")
print(dads_results$typical)

print("Stay-at-home dads: Target Hits")
print(dads_hits)

print("College undergraduates: Typical Levels")
print(college_results$typical)

print("College undergraduates: Target Hits")
print(college_hits)

print("Sophia: Typical Levels")
print(sophia_results$typical)

print("Sophia: Target Hits")
print(sophia_hits)

```

## Write up

STAY AT HOME DADS
Demographic Definition: Male, Age 25–50, Married or partnered, Not employed / primary caregiver, At least one child under 12
Key Activities: Childcare, Household chores, Sleep, Personal leisure, Exercise
Target levels: Childcare: 3 hours, Chores: 1.5 hours, Sleep: 8 hours, Leisure: 1 hour, Exercise: 30 minutes
% Meeting Targets: Childcare: 62%, Chores: 48%, Sleep: 36%, Leisure: 41%, Exercise: 19%

COLLEGE UNDERGRAD STUDENTS
Demographic Definition: Age 18–24, Enrolled full time, Living on campus or near campus, Mix of employed and unemployed students
Key Activities: Sleep, Studying / class time, Socializing, Screen time / leisure, Exercise
Target Levels: Sleep: 8 hours, Studying: 3 hours, Socializing: 1 hour, Screen time: ≤ 3 hours, Exercise: 30 minutes
% Meeting Targets: Sleep: 42%, Studying: 28%, Socializing: 55%, Screen time target: 31%, Exercise: 22%

SINGLE PROFESSIONAL WOMAN AGE 30-39
Demographic Definition: Female, Age 30–39, Unmarried, Employed full time, Lives independently
Key Activities: Sleep, Exercise, Socializing, Household chores, Work
Target Levels: Sleep: 7.5 hours, Exercise: 45 minutes, Socializing: 1 hour, Chores: 30 minutes, Work: 8.5 hours
% Meeting Targets: Sleep: 34%, Exercise: 18%, Socializing: 47%, Chores: 52%, Work: 49%



## AI Usage Statement

Copilot was used to help write the code in this project, but all non-code text was generated without the use of any Generative AI tools. Additionally, Copilot was used to brainstorm ideas for the final open-ended prompt.


Troubleshooting
```{r}

names(full_data)
unique(full_data$sex)
unique(full_data$marital_status)
unique(full_data$employment_status)


```


------------------------------------------------------------------------

This work ©2025 by laurenrichbetch was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #02](https://michael-weylandt.com/STA9750/mini/mini02.html)
