---
title: "STA 9750 laurenrichbetch MiniProject #**03**"
---


```{r}

#ACTIVITY 1

library(tidycensus)
library(tidyverse)
library(sf)

# 1. Ensure the data directory exists for caching
if (!dir.exists("data")) {
  dir.create("data")
}

# 2. Define the years, skipping 2020 as requested
years <- c(2015:2019, 2021:2024)

# 3. Retrieve ACS data
# We set tigris_use_cache to TRUE to handle shapefile caching
options(tigris_use_cache = TRUE)

pop_data_list <- map(years, ~ {
  message(paste("Downloading data for year:", .x))
  
  get_acs(
    geography = "state",
    variables = c(total_population = "B01003_001"),
    year = .x,
    survey = "acs1",
    geometry = TRUE,      # Downloads shapefiles
    cache_table = TRUE    # Caches variable metadata
  )
})

# 4. Combine into a tidy sf data frame
# We name the list elements by year, then bind them into one table
final_pop_df <- pop_data_list %>%
  set_names(years) %>%
  map_dfr(~.x, .id = "year") %>%
  mutate(year = as.numeric(year)) %>%
  rename(state = NAME, total_population = estimate) %>%
  select(GEOID, state, year, total_population, geometry)

# 5. Save the processed data to your data directory to avoid re-downloading
saveRDS(final_pop_df, "data/state_population_2015_2024.rds")

# Check the result
glimpse(final_pop_df)

```

```{r}

#ACTIVITY 2

library(tidyverse)
library(readxl)

# Helper function to clean Census numbers
clean_census_number <- function(x) {
  parse_number(as.character(x), na = c("N", "(X)", "Z", "NA", ""))
}

state_to_state_migration_2024 <- function() {
  dir_path <- "data/mp03"
  if (!dir.exists(dir_path)) dir.create(dir_path, recursive = TRUE)
  file_path <- file.path(dir_path, "migration_2024.xlsx")
  
  if (!file.exists(file_path)) {
    url <- "https://www2.census.gov/programs-surveys/demo/tables/geographic-mobility/2024/state-to-state-migration/State_to_State_Migration_Table_2024_T13.xlsx"
    download.file(url, file_path, mode = "wb")
  }
  
  # 1. Process 'Table'
  table_data <- read_excel(file_path, sheet = "Table", skip = 5, col_names = FALSE) %>%
    select(state_current = 1, state_1y = 2, population = 3) %>%
    mutate(population = clean_census_number(population)) %>%
    filter(!is.na(population), !is.na(state_current)) %>%
    mutate(year_current = 2024, year_1y = 2023)
  
  # 2. Process 'Supplemental'
  supp_data <- read_excel(file_path, sheet = "Supplemental - Current Res", skip = 5, col_names = FALSE) %>%
    select(state_current = 1, stayed_1 = 2, stayed_2 = 3) %>%
    mutate(
      stayed_1 = clean_census_number(stayed_1),
      stayed_2 = clean_census_number(stayed_2),
      population = coalesce(stayed_1, 0) + coalesce(stayed_2, 0)
    ) %>%
    filter(!is.na(state_current)) %>% 
    mutate(
      state_1y = state_current,
      year_current = 2024, year_1y = 2023
    ) %>%
    select(state_current, state_1y, population, year_current, year_1y)
  
  # 3. USE BIND_ROWS (This is the fix)
  final_data <- bind_rows(table_data, supp_data)
  
  return(final_data)
}

# --- Execution ---
my_migration_data <- state_to_state_migration_2024()

# --- Verification ---
glimpse(my_migration_data)

```

```{r}

#ACTIVITY 3

library(tidyverse)
library(readxl)

clean_census_number_2023 <- function(x) {
  parse_number(as.character(x), na = c("N", "(X)", "Z", "NA", ""))
}

state_to_state_migration_2023 <- function() {
  file_path <- "data/mp03/migration_2023.xlsx"
  states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])
  
  # 1. Stationary Data: Start at row 8 to skip all headers
  stayed_data <- read_excel(file_path, range = "A8:F58", col_names = FALSE) %>%
    select(state_current = 1, stayed_1 = 5, stayed_2 = 6) %>%
    mutate(
      population = clean_census_number_2023(stayed_1) + clean_census_number_2023(stayed_2),
      state_1y = state_current,
      year_current = 2023, year_1y = 2022
    ) %>%
    filter(!is.na(population)) # Drop any junk rows
  
  # 2. Flow Data: Start at row 8
  flow_data <- read_excel(file_path, range = "J8:DG58", 
                          col_names = vctrs::vec_interleave(states_with_dc, rep(NA, 51))) %>%
    select(any_of(states_with_dc)) %>%
    mutate(state_current = states_with_dc) %>%
    pivot_longer(cols = -state_current, names_to = "state_1y", values_to = "population") %>%
    mutate(population = clean_census_number_2023(population)) %>%
    filter(!is.na(population), state_current != state_1y) %>% # Drop headers and self-loops
    mutate(year_current = 2023, year_1y = 2022)
  
  return(bind_rows(stayed_data, flow_data))
}

# Run and verify
migration_data_2023 <- state_to_state_migration_2023()
summary(migration_data_2023$population)

```


```{r}

#ACTIVITY 4

# --- Task 4: Combined Function (Filtered) ---
state_to_state_migration <- function() {
  # 1. Fetch data from your two existing functions
  df_2024 <- state_to_state_migration_2024()
  df_2023 <- state_to_state_migration_2023()
  
  # 2. Combine and filter to keep only official states
  official_states <- c(state.name, "District of Columbia")
  
  combined_data <- bind_rows(df_2024, df_2023) %>%
    # Keep only records where BOTH origin and destination are in your official list
    filter(state_current %in% official_states,
           state_1y %in% official_states) %>%
    # Select only the columns needed for analysis
    select(state_current, state_1y, population, year_current, year_1y)
  
  return(combined_data)
}

# --- Execution ---
full_migration_data <- state_to_state_migration()

# --- Verification ---
# This will now return 51
print(paste("Unique states in current:", n_distinct(full_migration_data$state_current)))
print(paste("Unique states in 1y:", n_distinct(full_migration_data$state_1y)))

# This should show exactly 5,100 rows (51 states * 50 destinations * 2 years)
glimpse(full_migration_data)

```



```{r}

#ACTIVITY 5
metro_to_metro_migration <- function() {

    data_dir <- "data/mp03"
    if (!dir.exists(data_dir)) dir.create(data_dir, recursive = TRUE)

    api_url <- "https://api.census.gov/data/2020/acs/flows?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME&for=metropolitan%20statistical%20area/micropolitan%20statistical%20area:*"
    dest_file <- file.path(data_dir, "metro-to-metro-migration.json")
    
    if (!file.exists(dest_file)) {
        message("Fetching Metro Data from Census API...")
        # Ensure you have library(httr2) loaded
        httr2::request(api_url) |>
            httr2::req_perform() |>
            httr2::resp_body_string() |>
            writeLines(dest_file)
    }

    # FIX: Replace read_json with fromJSON
    raw_json <- jsonlite::fromJSON(dest_file, simplifyVector = TRUE)
    
    # Convert matrix to data frame and set headers from the first row
    metro_raw <- as.data.frame(raw_json[-1, ], stringsAsFactors = FALSE)
    colnames(metro_raw) <- raw_json[1, ]
    
    # Tidy the Data
    metro_clean <- metro_raw |>
        as_tibble() |>
        rename(metro_current = FULL1_NAME,
               metro_1y = FULL2_NAME,
               moved_in = MOVEDIN,
               moved_out = MOVEDOUT) |>
        mutate(across(c(moved_in, moved_out), as.integer)) |>
        select(metro_current, metro_1y, moved_in, moved_out) |>
        filter(!is.na(metro_current), !is.na(metro_1y))

    return(metro_clean)
}

metro_migration <- metro_to_metro_migration()

metro_migration <- metro_migration |>
    filter(str_detect(metro_1y, "Metro Area$|Micro Area$")) |>
    filter(!str_detect(metro_1y, "Outside Metro Area"))
```

```{r}
# JUST LOAD THE LIBRARIES - DO NOT INSTALL THEM HERE
library(dplyr)
library(knitr)
library(kableExtra)
library(scales)
library(usmap)
library(ggplot2)
library(jsonlite)
library(httr2)
library(stringr)

```

```{r}

#ACTIVITY 6 - QUESTION 1
#Which states have had the highest net population growth rates over the past decade?

install.packages("kableExtra")

library(dplyr)
library(tidyr)
library(sf)
library(scales)
library(knitr)
library(kableExtra) # Essential for borders and titles

# 1. Prepare and format the data
formatted_table <- final_pop_df %>%
  st_drop_geometry() %>% 
  filter(year %in% c(2015, 2024)) %>%
  select(state, year, total_population) %>%
  pivot_wider(names_from = year, values_from = total_population) %>%
  mutate(
    # Store raw numbers for sorting, but format for display
    raw_delta = (`2024` - `2015`) / `2015`,
    delta_pct = paste0(round(raw_delta * 100, 2), "%"),
    `2015` = comma(`2015`),
    `2024` = comma(`2024`)
  ) %>%
  arrange(desc(raw_delta)) %>%
  select(state, `2015`, `2024`, `Delta` = delta_pct)

# 2. Render the formatted table
formatted_table %>%
  kable(caption = "Fastest Growing States for 2015 to 2024", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = F)

```

## WRITE UP
Idaho, Utah Florida and Texas had the highest net population growth rates over the past decade. This could be due to the affordability and quality of life that the states have to offer. The cost of living is comparably lower to other states, and the states are also known for the outdoor activities. Outdoor activities have likely become more important due to hybrid and remote work, which has freed up leisure time versus pre-pandemic.

```{r}

#ACTIVITY 6 - QUESTION 2
#If you meet someone who moved to New York State in the last year, what are the most likely states they moved from? PART 1

library(dplyr)
library(ggplot2)
library(scales)

# 1. Filter for people who moved into New York State (Numeric Data Only)
ny_inbound_plot_data <- full_migration_data |>
  # Keep only records moving TO New York
  filter(state_current == "New York") |>
  # Exclude those who stayed (moved from NY to NY)
  filter(state_1y != "New York") |>
  # Aggregate by origin state
  group_by(state_1y) |>
  summarize(total_movers = sum(population, na.rm = TRUE)) |>
  # Sort to find the top sources
  arrange(desc(total_movers)) |>
  slice_head(n = 10)

# 2. Render the Bar Chart
ggplot(ny_inbound_plot_data, aes(x = reorder(state_1y, total_movers), y = total_movers)) +
  geom_col(fill = "steelblue") +
  coord_flip() + # Makes the state names readable
  labs(
    title = "Top 10 States People Moved From to New York (2023-2024)",
    x = "Origin State",
    y = "Number of Movers"
  ) +
  # Add commas to the X-axis (the number axis)
  scale_y_continuous(labels = comma) +
  theme_minimal()

```

## WRITE UP
California stands out at the top state people move from to New York. This could be due to many factors including wanting to experience all four seasons given most parts of California are warm and do not get snow. New York also has the best work opportunities within Finance and Law which could be appealing. New York and California are comparable in cost of living, therefore, people in California are some of the few who could move to New York and maintain their lifestyle.

```{r}

#ACTIVITY 6 - QUESTION 2
#If one of your friends announces they are moving out of the state, where are they most likely to move? PART 2

library(dplyr)
library(ggplot2)
library(scales)

# 1. Prepare the data (Numeric Only)
ny_outbound_plot_data <- full_migration_data |>
  # Keep only records where New York was the origin state
  filter(state_1y == "New York") |>
  # Exclude those who stayed (moved from NY to NY)
  filter(state_current != "New York") |>
  # Aggregate by destination state
  group_by(state_current) |>
  summarize(total_movers = sum(population, na.rm = TRUE)) |>
  # Sort to find the top destinations
  arrange(desc(total_movers)) |>
  slice_head(n = 10)

# 2. Render the Bar Chart
ggplot(ny_outbound_plot_data, aes(x = reorder(state_current, total_movers), y = total_movers)) +
  geom_col(fill = "skyblue") + # Using a different color for outbound
  coord_flip() + # Makes the state names readable
  labs(
    title = "Top 10 States People Moved TO from New York (2023-2024)",
    x = "Destination State",
    y = "Number of Movers"
  ) +
  # Add commas to the axis numbers automatically
  scale_y_continuous(labels = comma) +
  theme_minimal()



```

## WRITE UP
Arizona is the clear choice for people to move to from New York. The primary driver of this trend is likely the cost of living and housing affordability. New York is known to have some of the highest housing costs, therefore, people who are looking to have a family, or spend their money in other areas including travel would find Arizona an appealing place to be. In addition, Arizona has nice weather and could be appealing for people who want to experience less cold weather, and more sunshine.

```{r}

#ACTIVITY 6  - QUESTION 3
#If you meet someone who moved to New York City in the last year, what are the most likely metro areas they moved from? PART 1

# Load all required libraries
library(dplyr)
library(ggplot2)
library(scales)
library(jsonlite)
library(httr2)
library(stringr)

# 3. Process the data for NYC
nyc_inbound_plot_data <- metro_migration |>
  filter(metro_current == "New York-Newark-Jersey City, NY-NJ-PA Metro Area") |>
  filter(metro_1y != metro_current) |>
  group_by(metro_1y) |>
  summarize(total_movers = sum(moved_in, na.rm = TRUE)) |>
  arrange(desc(total_movers)) |>
  slice_head(n = 5)

# 4. Render the Donut Chart
ggplot(nyc_inbound_plot_data, aes(x = 2, y = total_movers, fill = metro_1y)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void() + 
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7), 
    legend.title = element_text(size = 9)
  ) +
  labs(
    title = "Top 5 Metro Areas Moving INTO NYC",
    fill = "Origin Metro"
  ) +
  guides(fill = guide_legend(ncol = 2))


```

## WRITE UP
Four of the top five metro areas moving into NY are on the east coast. This trend could be driven by college students who went to school in nearby cities and want to start their careers in NY. Generally speaking, NY offers a more diverse experience versus the other cities - including dining and theater, which could be appealing for people who are looking to have a wide variety of cultural experiences.


```{r}

#ACTIVITY 6 - QUESTION 3
#If one of your friends announces they are moving out of the city, where are they most likely to move? PART 2

library(dplyr)
library(ggplot2)
library(scales)

# 1. Process data (Keep total_movers numeric for plotting)
nyc_outbound_plot_data <- metro_migration |>
  # Note: Based on your logic, we filter for NYC as the origin (metro_current)
  filter(metro_current == "New York-Newark-Jersey City, NY-NJ-PA Metro Area") |>
  filter(metro_1y != metro_current) |>
  group_by(metro_1y) |>
  summarize(total_movers = sum(moved_out, na.rm = TRUE)) |>
  arrange(desc(total_movers)) |>
  slice_head(n = 5)

# 2. Render the Donut Chart
ggplot(nyc_outbound_plot_data, aes(x = 2, y = total_movers, fill = metro_1y)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  theme_void() + 
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 7), 
    legend.title = element_text(size = 9)
  ) +
  labs(
    title = "Top 5 Metro Areas People Moved TO from NYC",
    fill = "Destination Metro"
  ) +
  # Force legend into 2 columns for a compact look
  guides(fill = guide_legend(ncol = 2))

```

## WRITE UP
Four of the top five metro areas moving into NY are on the east coast. People who are leaving NY for other east coast cities are likely looking for a more affordable option. People who move to Los Angeles are clearly not looking for affordability - and could instead be looking to escape the east coast weather. Likewise, the retired community is likely driving the trend of moving from NY to Miami in search of warmer weather.

```{r}

#ACTIVITY 6 - QUESTION 4
##Which states have had the highest amounts of in-migration and out-migration? Furthermore, which states have had the highest amount of net migration?

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

# 1. Calculate In-Migration (State is destination)
inbound <- full_migration_data |>
  filter(state_current != state_1y) |>
  group_by(state_current) |>
  summarize(total_in = sum(population, na.rm = TRUE)) |>
  rename(state = state_current)

# 2. Calculate Out-Migration (State is origin)
outbound <- full_migration_data |>
  filter(state_current != state_1y) |>
  group_by(state_1y) |>
  summarize(total_out = sum(population, na.rm = TRUE)) |>
  rename(state = state_1y)

# 3. Join them into one summary
state_migration_summary <- inbound |>
  full_join(outbound, by = "state") |>
  mutate(
    total_in = replace_na(total_in, 0),
    total_out = replace_na(total_out, 0),
    net_migration = total_in - total_out
  ) |>
  arrange(desc(net_migration))

# 4. Display the Top 10
state_migration_summary |>
  slice_head(n = 10) |>
  mutate(across(c(total_in, total_out, net_migration), comma)) |>
  kable(
    col.names = c("State", "Total In-Migration", "Total Out-Migration", "Net Migration"),
    caption = "Top 10 States by Net Migration",
    align = "lrrr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = F)
```

## WRITE UP
Similar to earlier findings, Arizona is the clear choice for people to move to from other states, including New York. This is likely due to the sunlight and mild winters, where the sun shines in Arizona for about 300 days each year! The climate is also dry which is most often times preferred to humidity. Likewise, the great weather lends itself to year-round outdoor activities including golf, cycling and tennis.

```{r}

#ACTIVITY 6 - QUESTION 4 - HEAT MAP

library(usmap)
library(ggplot2)
library(scales)
library(dplyr)

# 1. Prepare your summary data
# Note: Ensure the 'state' column contains full state names (e.g., "Texas", "Florida")
# as usmap can automatically map these.
map_data_full <- state_migration_summary 

# 2. Render the Map
plot_usmap(data = map_data_full, values = "net_migration", color = "white") +
  scale_fill_gradientn(
    colors = c("#D73027", "#FEE08B", "#1A9850"), # Red (Loss) to Yellow (Zero) to Green (Gain)
    name = "Net Migration",
    label = scales::comma,
    # This centers the color scale at zero so neutral states are yellow
    values = scales::rescale(c(min(map_data_full$net_migration, na.rm = TRUE), 
                               0, 
                               max(map_data_full$net_migration, na.rm = TRUE)))
  ) +
  labs(
    title = "Net Migration by State",
    subtitle = "Positive (Green) indicates inflow; Negative (Red) indicates outflow"
  ) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 10)
  )

```


```{r}

#ACTIVITY 6  - QUESTION 5
##Which metro areas have had the highest amounts of total in-migration and out-migration? Furthermore, which metro areas have had the highest amount of net migration?

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

# 1. Summarize migration by Metro Area
metro_migration_summary <- metro_migration |>
  # Exclude self-loops (moves within the same metro) if they exist
  filter(metro_current != metro_1y) |>
  group_by(metro_current) |>
  summarize(
    total_in = sum(moved_in, na.rm = TRUE),
    total_out = sum(moved_out, na.rm = TRUE)
  ) |>
  mutate(net_migration = total_in - total_out) |>
  arrange(desc(net_migration))

# 2. Display the top 10 Metro Areas by Net Migration
metro_migration_summary |>
  select(metro_current, total_in, total_out, net_migration) |>
  slice_head(n = 10) |>
  mutate(across(c(total_in, total_out, net_migration), comma)) |>
  kable(
    col.names = c("Metro Area", "Total In-Migration", "Total Out-Migration", "Net Migration"),
    caption = "Top 10 Metro Areas by Net Migration",
    align = "lrrr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = F)

```
## WRITE UP
Metro areas that have the most in-migration share key similarities. First and foremost, several of the states do not have personal income tax (Florida, Texas and Tennessee), while Arizona has a low tax structure - this is appealing for people who are tired of paying taxes! To no surprise, these Metro areas are all relatively warm.

```{r}

# ACTIVITY 6 - QUESTION 6
# Which states have the highest fraction of residents who lived there last year?

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

# Calculate the fraction of residents who stayed
retention_rates <- full_migration_data |>
  # 1. Group by state
  group_by(state_current) |>
  # 2. Calculate stayed population vs total population
  summarize(
    total_pop = sum(population, na.rm = TRUE),
    stayed_pop = sum(population[state_current == state_1y], na.rm = TRUE),
    # Calculate the fraction
    retention_fraction = stayed_pop / total_pop
  ) |>
  # 3. Sort by highest retention (fraction)
  arrange(desc(retention_fraction))

# Display the top 10 states with comma formatting
retention_rates |>
  mutate(
    total_pop = comma(total_pop),            # Add commas
    stayed_pop = comma(stayed_pop),          # Add commas
    retention_fraction = percent(retention_fraction, accuracy = 0.1)
  ) |>
  slice_head(n = 10) |>
  kable(
    col.names = c("State", "Total Population", "Stayed Population", "Retention Fraction"),
    caption = "Top 10 States with Highest Fraction of Residents Who Stayed",
    align = "lllr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = F)

```

## WRITE UP
The top 10 states where people have stayed have very similar retention amounts where each state has it's own reason why people would want to stay. To name a new, California and Texas have warm weather while New Jersey is known for it's public school system.

```{r}

# ACTIVITY 6 - QUESTION 7
# Which state has the highest fraction of its population growth attributable to net internal migration?

library(dplyr)
library(scales)
library(knitr)
library(kableExtra)

# 1. Load your population growth data
state_pop_data <- read_rds("./data/state_population_2015_2024.rds")

# 2. Calculate Total Growth (2023 to 2024)
state_growth_data <- state_pop_data |>
  filter(year %in% c(2023, 2024)) |>
  group_by(state) |>
  summarize(total_growth = total_population[year == 2024] - total_population[year == 2023])

# 3. Calculate Net Migration from 'full_migration_data'
inbound <- full_migration_data |>
  filter(state_current != state_1y) |>
  group_by(state_current) |>
  summarize(total_in = sum(population, na.rm = TRUE))

outbound <- full_migration_data |>
  filter(state_current != state_1y) |>
  group_by(state_1y) |>
  summarize(total_out = sum(population, na.rm = TRUE))

net_migration_data <- inbound |>
  full_join(outbound, by = c("state_current" = "state_1y")) |>
  mutate(
    total_in = replace_na(total_in, 0),
    total_out = replace_na(total_out, 0),
    net_migration = total_in - total_out
  )

# 4. Join, calculate fraction, and format
final_analysis <- net_migration_data |>
  inner_join(state_growth_data, by = c("state_current" = "state")) |>
  mutate(
    migration_fraction = net_migration / total_growth
  ) |>
  filter(total_growth > 0) |> # Only include states with positive growth
  arrange(desc(migration_fraction))

# 5. Display the table with comma-formatted integers and percentages
final_analysis |>
  select(state_current, total_growth, net_migration, migration_fraction) |>
  mutate(
    total_growth = comma(total_growth),
    net_migration = comma(net_migration),
    migration_fraction = percent(migration_fraction, accuracy = 0.1, big.mark = ",")
  ) |>
  slice_head(n = 10) |>
  kable(
    col.names = c("State", "Total Growth", "Net Migration", "Migration Fraction"),
    caption = "Top 10 States: Fraction of Growth from Net Internal Migration",
    align = "lrrr"
  ) |>
  kable_styling(bootstrap_options = c("striped", "hover", "bordered"), full_width = F)

```

## WRITE UP
Wyoming, Arizona and Mississippi heavily depend on migration into the state versus new births. These states must attract residents of other states in order to sustain economic and social health - if the migration was not net positive - then these states would experience a shriking labor force, and tax base.

```{r}

# ACTIVITY 6 - QUESTION 8
# Which state had the highest amount of migration into your state? Of the people who moved out of your state, what was the most popular destination state?

library(dplyr)
library(knitr)
library(kableExtra)
library(scales)

# 1. Define your state
my_state <- "Texas"

# 2. Who moved INTO Texas?
# We look for where state_current is Texas (and they came from elsewhere)
into_my_state <- full_migration_data |>
  filter(state_current == my_state, state_1y != my_state) |>
  arrange(desc(population)) |>
  slice_head(n = 5) |>
  mutate(population = comma(population))

# 3. Where did people go when moving OUT of Texas?
# We look for where state_1y is Texas (and they went elsewhere)
out_of_my_state <- full_migration_data |>
  filter(state_1y == my_state, state_current != my_state) |>
  arrange(desc(population)) |>
  slice_head(n = 5) |>
  mutate(population = comma(population))

# 4. Display the results
kable(into_my_state %>% select(state_1y, population), 
      col.names = c("Origin State", "Movers into TX"),
      caption = "Top 5 States Moving INTO Texas") |> 
  kable_styling(full_width = F)

kable(out_of_my_state %>% select(state_current, population), 
      col.names = c("Destination State", "Movers from TX"),
      caption = "Top 5 Destination States for Texans Moving Out") |> 
  kable_styling(full_width = F)


```


## WRITE UP
In line with national trends, people who are moving out of Texas are migrating to Arizona for reasons discussed earlier in this project.

```{r}

library(usmap)
library(ggplot2)
library(dplyr)
library(scales)

# 1. Prepare data for ALL states moving INTO Texas
tx_inbound_all <- full_migration_data |>
  filter(state_current == "Texas", state_1y != "Texas") |>
  group_by(state_1y) |>
  summarize(movers = sum(population, na.rm = TRUE)) |>
  rename(state = state_1y)

# 2. Render the heat map for all states
plot_usmap(data = tx_inbound_all, values = "movers", color = "white") +
  scale_fill_gradient(
    low = "#ADD8E6",  # Light Baby Blue
    high = "#0047AB", # Darker Baby Blue
    name = "Movers to TX",
    label = scales::comma,
    na.value = "grey90"
  ) +
  labs(
    title = "Migration INTO Texas from All States",
    subtitle = "Higher intensity represents more people moving to Texas"
  ) +
  theme(legend.position = "right")

```
## WRITE UP
California and Florida are the top states where people are moving from into Texas. People moving out of California are likely looking to maintain a warmer climate, but could be looking for a more affordable option. For people migrating out of Florida, they could be looking to escape communities that skew older as Florida is a popular place for people who are retired to move.

```{r}

# ACTIVITY 6 - QUESTION 9
# Which metro area, not located in your state, has the highest amount of migration into your state’s largest metro? To which metro area outside of your state do the most residents of your largest metro area move?

library(dplyr)
library(stringr)
library(scales)
library(knitr)
library(kableExtra)

# 1. Define your metro (ensure this matches the exact spelling in your data!)
# Note: Based on your sample, it is "Austin-Round Rock-GeorgeTown, TX Metro Area"
my_metro <- "Austin-Round Rock-GeorgeTown, TX Metro Area"

# 2. Who is moving INTO Austin (from outside Texas)?
# We filter for rows where 'metro_current' is Austin
# AND 'metro_1y' does not contain "TX"
into_austin <- metro_migration |>
  filter(metro_current == my_metro) |>
  filter(!str_detect(metro_1y, "TX")) |> # Filters out any metro with TX in the name
  group_by(metro_1y) |>
  summarize(total_in = sum(moved_in, na.rm = TRUE)) |>
  arrange(desc(total_in)) |>
  slice_head(n = 10) |>
  mutate(total_in = comma(total_in))

# 3. Where are people moving TO from Austin (outside Texas)?
# We filter for rows where 'metro_1y' is Austin
# AND 'metro_current' does not contain "TX"
out_of_austin <- metro_migration |>
  filter(metro_1y == my_metro) |>
  filter(!str_detect(metro_current, "TX")) |>
  group_by(metro_current) |>
  summarize(total_out = sum(moved_out, na.rm = TRUE)) |>
  arrange(desc(total_out)) |>
  slice_head(n = 10) |>
  mutate(total_out = comma(total_out))

# 4. Display the results
kable(into_austin, col.names = c("Origin Metro", "In-Migrants"), 
      caption = "Top 10 Metros Moving INTO Austin (Outside TX)") |> 
  kable_styling(bootstrap_options = "striped")

kable(out_of_austin, col.names = c("Destination Metro", "Out-Migrants"), 
      caption = "Top 10 Metros Moving OUT OF Austin (Outside TX)") |> 
  kable_styling(bootstrap_options = "striped")

```

## WRITE UP
People who are moving out of Austin, TX are largely migrating to California cities (four of of the top 10), sticking with the theme of nice weather - people moving out of a sunny climate are likely not willing to move somewhere that has harsh winters, though some are willing to make the sacrifice for the cultural experience of living in New York-Newark-Jersey City.


```{r}
# ACTIVITY 6 - QUESTION 10
# Are there any metro areas that have a particularly high connection to your state’s largest metro?

library(dplyr)
library(scales)
library(knitr)
library(kableExtra)

# 1. Define your metro
my_metro <- "Austin-Round Rock-GeorgeTown, TX Metro Area"

# 2. Calculate In-Migration Shares (Who moves INTO Austin from outside TX?)
# First, calculate the total pool of inbound movers from non-TX metros
in_total <- metro_migration |>
  filter(metro_current == my_metro, !str_detect(metro_1y, "TX")) |>
  summarize(total = sum(moved_in, na.rm = TRUE)) |>
  pull(total)

inbound_shares <- metro_migration |>
  filter(metro_current == my_metro, !str_detect(metro_1y, "TX")) |>
  group_by(metro_1y) |>
  summarize(movers = sum(moved_in, na.rm = TRUE)) |>
  mutate(share = movers / in_total) |>
  arrange(desc(share)) |>
  slice_head(n = 10) |>
  mutate(movers = comma(movers),
         share = percent(share, accuracy = 0.1))

# 3. Calculate Out-Migration Shares (Who moves OUT of Austin to outside TX?)
out_total <- metro_migration |>
  filter(metro_1y == my_metro, !str_detect(metro_current, "TX")) |>
  summarize(total = sum(moved_out, na.rm = TRUE)) |>
  pull(total)

outbound_shares <- metro_migration |>
  filter(metro_1y == my_metro, !str_detect(metro_current, "TX")) |>
  group_by(metro_current) |>
  summarize(movers = sum(moved_out, na.rm = TRUE)) |>
  mutate(share = movers / out_total) |>
  arrange(desc(share)) |>
  slice_head(n = 10) |>
  mutate(movers = comma(movers),
         share = percent(share, accuracy = 0.1))

# 4. Display the results
kable(inbound_shares, col.names = c("Origin Metro", "Movers", "Share of Total Inbound"), 
      caption = "Top 10 Metros Moving INTO Austin (Outside TX) by Share") |> 
  kable_styling(bootstrap_options = "striped")

kable(outbound_shares, col.names = c("Destination Metro", "Movers", "Share of Total Outbound"), 
      caption = "Top 10 Metros Moving OUT OF Austin (Outside TX) by Share") |> 
  kable_styling(bootstrap_options = "striped")

```

## WRITE UP
All metros individually contribute less than 10% of total migration to Austin, Texas indicating this geo area has a wide appeal where interest is not isloated to any given location.


```{r}

# ACTIVITY 7 - PART 1
library(dplyr)
library(tidyr)
library(sf)
library(scales)
library(knitr)
library(kableExtra)

# 1. DEFINE RENAMED FUNCTION
my_pop_projection_2030 <- function(current_pop, growth, migration_df) {
  natural_growth <- current_pop * (1 + growth)
  
  net_migration <- sapply(names(current_pop), function(s) {
    incoming <- sum(migration_df$m_ij[migration_df$state_current == s] * current_pop[migration_df$state_1y[migration_df$state_current == s]], na.rm = TRUE)
    outgoing <- sum(migration_df$m_ij[migration_df$state_1y == s] * current_pop[s], na.rm = TRUE)
    return(incoming - outgoing)
  })
  
  return(natural_growth + net_migration)
}

# 2. PREPARE POPULATIONS
state_pops <- final_pop_df |>
  st_drop_geometry() |> 
  filter(year %in% c(2022, 2023, 2024)) |> 
  select(state, year, total_population) |>
  pivot_wider(names_from = year, values_from = total_population)

# 3. FIT PARAMETERS
fit_model <- function(t0, t1) {
  P_t0 <- sum(state_pops[[as.character(t0)]], na.rm = TRUE)
  P_t1 <- sum(state_pops[[as.character(t1)]], na.rm = TRUE)
  g <- (P_t1 - P_t0) / P_t0
  
  mig_rates <- full_migration_data |>
    filter(year_current == t1, year_1y == t0) |>
    filter(state_current != state_1y) |>
    inner_join(state_pops, by = c("state_current" = "state")) |>
    rename(pop_i = as.character(t0)) |>
    inner_join(state_pops, by = c("state_1y" = "state")) |>
    rename(pop_j = as.character(t0)) |>
    mutate(m_ij = population / sqrt(pop_i * pop_j))
  
  return(list(g = g, m = mig_rates))
}

fit1 <- fit_model(2022, 2023)
fit2 <- fit_model(2023, 2024)
avg_g <- (fit1$g + fit2$g) / 2
avg_m <- bind_rows(fit1$m, fit2$m) |>
  group_by(state_current, state_1y) |>
  summarize(m_ij = mean(m_ij, na.rm = TRUE), .groups = 'drop')

# 4. RUN PROJECTION
current_pop <- setNames(state_pops$`2024`, state_pops$state)

# Using the RENAMED function
for (y in 2025:2030) {
  current_pop <- unname(my_pop_projection_2030(current_pop, avg_g, avg_m))
  names(current_pop) <- state_pops$state
}

# 5. CLEAN DISPLAY
final_df <- tibble(
  State = names(current_pop),
  `Projected 2030 Population` = as.numeric(current_pop)
) |>
  mutate(`Projected 2030 Population` = comma(round(`Projected 2030 Population`))) |>
  arrange(desc(as.numeric(gsub(",", "", `Projected 2030 Population`))))

final_df |>
  head(10) |>
  kable(caption = "Projected 2030 Populations (Top 10)") |>
  kable_styling(bootstrap_options = "striped")

```


```{r}

#ACTIVITY 8


library(tidyverse)
library(gt)

# 1. PREPARE 2030 DATA
# We use the 'current_pop' vector from the previous activity
apportionment_2030_data <- tibble(
  state = names(current_pop),
  population = as.numeric(current_pop)
)

# 2. APPLY HUNTINGTON-HILL
# We assign seats 1 through 435 based on the highest priority scores
N_DISTRICTS <- 435
MAX_DISTRICTS <- 100 

apportionment_2030_results <- apportionment_2030_data |>
  expand_grid(cd = seq(0, MAX_DISTRICTS)) |>
  mutate(
    hh_den = sqrt(cd * (cd + 1)),
    hh_weight = population / hh_den
  ) |>
  # Select the 435 seats with the highest priority scores
  slice_max(hh_weight, n = N_DISTRICTS) |>
  group_by(state) |>
  summarize(
    population = first(population),
    n_districts = n()
  ) |>
  mutate(pop_per_district = population / n_districts) |>
  arrange(desc(n_districts), desc(population), state)

# 3. RENDER THE GT TABLE
apportionment_2030_results |>
  gt(id = "tbl_2030_apportionment", rowname_col = "state") |>
  tab_header(title = "Projected 2030 Congressional Apportionment") |>
  cols_label(
    state = md("State"),
    population = "Projected 2030 Population",
    n_districts = "Number of Seats",
    pop_per_district = "Approx. Pop per District"
  ) |>
  fmt_integer(c(population, pop_per_district)) |>
  data_color(columns = pop_per_district, palette = "PuOr") |>
  grand_summary_rows(
    columns = c(population, n_districts),
    fns = list("Total" = ~ sum(.)),
    fmt = ~ fmt_integer(.)
  ) |>
  tab_source_note(md("Apportionments computed using the [Huntington-Hill Method](https://en.wikipedia.org/wiki/Huntington%E2%80%93Hill_method)."))

```

## ACTIVITY 9 - WRITE UP


GEO TARGETING:
•	In-Migration Targets: target Los Angeles-Long Beach-Anaheim, CA and Miami-Fort Lauderdale-West Palm Beach, FL. It’s expensive to live in these locations, and the populations are dense. The campaign will showcase that Texas is affordable, has no state personal income tax and has job opportunities in tech and energy.
•	Out-Migration Targets: Target Phoenix-Mesa-Scottsdale, AZ and Denver-Aurora-Lakewood, CO. These are key areas of out-migration from Texas, so we will try to convince people to come back home. The campaign messaging will highlight career opportunities in Texas, and quality of life benefits that people will not get in those other areas.
•	Congressional Targets: Place advertising in Kansas City, MO-KS and St. Louis, MO-IL. Missouri could lose a congressional seat given the population.

CAMPAIGN PARAMETERS AND EXPECTED IMPACT:
•	On average, a state needs an increase of 500,000-700,000 residents to secure a congressional seat.
•	We will need $25MM budget for a multi-channel campaign that will include Linear TV, OOH and all Digital channels.
•	We are hoping to gain an additional 150,000 people over the next three years which defending the base.
MARKETING VERBIAGE:

•	Slogan: Your Future is Bigger in Texas
•	If you are tired of all your money going to taxes, Texas is the place for you! Come build your career in tech, and spend your money on what makes you the happiest in life.


::: {.callout-note title="AI Usage Statement" collapse="true"}

Gemini was used to help write the code in this project, but all non-code text was generated without the use of any Generative AI tools. Additionally, Gemini was used to provide additional background information on the topic and to brainstorm ideas for the final open-ended prompt.

:::
